Fail Faster

### 2025-03-16 paper/paper.tex
staging dsls
staging background

Given these strict timelines, it is important to think of PBT as
performance-sensitive process. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (DSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning. \jwc{... to find bugs fast}

@@

@@

\jwc{
  \begin{itemize}
    \item In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
with each stage producing the code to be run in the next.
\item This is an old idea, with roots going back to quasiquotation in LISP, and picking up steam again in the 1990s with MetaML \cn{}.
\item For the purposes of this paper, we will only consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
which produces more code, which is then compiled and run at run time.
\item Staging has many applications, but chief among them is its use for \emph{optimization}. We can write programs such that during the compile time stage, they are partially evaluated to eliminate
abstraction overhead. 
\item Many languages have some degree of staging functionality, either provided as a library \cn or build directly into the language's implementation \cn.
\item For this paper, we use MetaOCaml \cn for OCaml staging, but all of the staging concepts are portable to any other language with multi-stage programming functionality.
  \end{itemize}
}

\jwc{This section is going to need work.}

MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
value of type \texttt{t code} at compile time is a (potentially open) OCaml term
of type \texttt{t}.

@@

\subsection{Monadic DSL Abstraction Overhead}
\jwc{
  \begin{itemize}
    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

>>>>

Given these strict timelines, it is important to think of PBT as
performance-sensitive process. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning. \jwc{... to find bugs fast}

@@

\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}

@@

In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
with each stage producing the code to be run in the next.
This is an old idea, with roots going back to quasiquotation in LISP \cn{}, and picking up steam again in the 1990s with MetaML \cn{}.
For the purposes of this paper, we will primarily consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
which produces more code, which is then compiled and run at run time.

\subsubsection{Staging in eDSL Construction}

One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
In the literature, this effect is known as ``abstraction overhead'': the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cn, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cn, Haskell \cn) to experimental compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}, Kotlin \cn).
Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}

For this paper, we have implemented staged PBT libraries in both OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is written in OCaml.
However, we believe all of the concepts presented are portable across languages with staging support, and we have reimplemented \name in Scala 3.

\subsubsection{Staging in MetaOCaml}
MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
value of type \texttt{t code} at compile time is a (potentially open) OCaml term
of type \texttt{t}.

@@

\subsection{Monadic DSL Abstraction Overhead}

\jwc{
  \begin{itemize}
    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}
